---
title: "UMFPACK Interace for R"
author: "Rob van Harrevelt"
date: "`r Sys.Date()`"
output: pdf_document
linkcolor: blue
vignette: >
  %\VignetteIndexEntry{UMFPACK Interface for R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE, comment = "")
library(umfpackr)
```

# Introduction

Package `umfpackr` provides an interface to UMFPACK, a sparse direct solver 
written in  C. The package `umfpackr` contains the functions
`umf_solve` and `umf_solve_nl` for solving 
linear and non-linear systems of equations, respectively.
This vignette describes the UMPFACK interace implemented in
`umfpackr` in detail.

# System dependencies

UMFPACK is part of  [SuiteSparse](http://faculty.cse.tamu.edu/davis/suitesparse.html),
a suite of libraries for sparse linear algebra.

On Linux and other non-Windows platforms, `umfpackr` assumes that
a full installation of SuiteSparse is available. On Ubuntu 18.04 installation
is very easy:
```
sudo apt-get install libsuitesparse-dev
```
Something similar is probably possible for other Linux platforms. Otherwise
you have to build SuiteSparse from source.

On Windows, it difficult to install the full SuiteSparse package and use
it in R. However, it is possible to use the UMFPACK and AMD libraries
in an R package. These libraries are sufficient to 
solve sparse linear
systems using the AMD ordering method. For the Windows
implementation of `umfpackfr`,  the souce of these libraries is included
in the package and it is therefore not necesarry to install SuiteSparse on Windows. 
The drawback is that it is not possible to use the METIS
ordering method on Windows. The METIS method can handle larger
matrices than the standard AMD method.

# Ordering methods

UMFPACK can use several ordering methods. As explained in the previous section,
on Windows only one ordering method is possible (AMD), but on other platforms
the user can select several other ordering methods.

In R package `umfpackr`, the following ordering methods can be chosen:

  * AMD. This is the default. AMD is used for the symmetric strategy and 
  COLAMD for the unsymmetric strategy. 
  * CHOLMOD. It first tries
AMD or COLAMD (depending on what strategy is used). If that method gives low
fill-in, it is used without trying METIS at all. Otherwise METIS is tried, and the ordering
(AMD/COLAMD or METIS) giving the lowest fill-in is used.
  * METIS. Use METIS.
  * BEST. Try three methods (AMD/COLAMD, METIS and NESDIS) and take the best. 
  NESDIS is CHOLMOD’s nested dissection ordering, based on METIS and CCAMD/CCOLAMD. 
  This results the highest analysis time, but the lowest numerical factorization time.
  
  The AMD method is usually slightly faster than the other ordering methods,
  but METIS can handle larger matrices than AMD. Therefore is sometimes
  necessary to the the METIS ordering (in that case METIS ordering is usually
  selected when CHOLMOD ordering is used).
  
More details can be found in the [_UMFPACK User Guide_](../doc_umfpack/UMFPACK_UserGuide.PDF) 
and references therein.
  
In `umfpackr` the ordering method can be selected by specifying 
argument `umf_control` of function `umf_solve` or `umf_solve_nl`.
This argument should be a named list. To select  METIS
ordering, specify `umf_control = list(ordering = "METIS")`.
An example were the CHOLMOD  is chosen on Linux and AMD on Windows:

```{r}
dslnex <- function(x) {
   y <- numeric(2)
   y[1] <- x[1]^2 + x[2]^2 - 2
   y[2] <- exp(x[1]-1) + x[2]^3 - 2
   return(y)
}

jacdsln <- function(x) {
   n <- length(x)
   Df <- matrix(numeric(n*n),n,n)
   Df[1,1] <- 2*x[1]
   Df[1,2] <- 2*x[2]
   Df[2,1] <- exp(x[1]-1)
   Df[2,2] <- 3*x[2]^2
   return(as(Df, "dgCMatrix"))
}

xstart <- c(2,3)

# use CHOLMOD ordering on Linux and AMD on Windows
ord <- if (.Platform$OS.type != "windows")  "CHOLMOD" else "AMD"
umf_solve_nl(xstart, dslnex, jacdsln, umf_control = list(ordering = ord))
```
